Another Look at Regularized Discriminant Analysis for High-Dimensional Classification

### 2013-04-17 reports/paper-rda.Rnw
Renamed GRDA classifier to HDRDA. Also, applied edits from Dean.

results$lambda <- as.numeric(results$lambda)
results$gamma <- as.numeric(results$gamma)
results$Dudoit <- as.numeric(results$Dudoit)
results$GRDA <- as.numeric(results$GRDA)
results$Guo <- as.numeric(results$Guo)
results$Pang <- as.numeric(results$Pang)
results$Tong <- as.numeric(results$Tong)
results$Witten <- as.numeric(results$Witten)

@@

# Summary of the optimal tuning-parameter estimates for each data set and number
# of variables selected, d.
grda_optimal <- subset(results, select = c(data_set, d, lambda, gamma))
grda_optimal$lambda <- factor(grda_optimal$lambda, levels = seq_lambda)
grda_optimal$gamma <- factor(grda_optimal$gamma, levels = seq_gamma)

grda_optimal <- ddply(grda_optimal, .(data_set, d), function(x) {
  x <- subset(x, select = c(lambda, gamma))
  melt(table(x))
})

@@

A reliable supervised classification of tumors from gene expression microarray
data is essential for successful diagnosis and treatment of diseases such as
cancer, but the analysis of microarray data often presents difficult challenges
because the number of gene expression levels is typically much greater than the
sample size. Consequently, several sparse and regularized classifiers have been
proposed, often with strict covariance assumptions or eigenvalue adjustments to
ensure that the classifiers are well-posed. Here, we consider an improved
version of the regularized discriminant analysis (\emph{RDA}) classifier
popularized by \cite{Friedman:1989tm}, which requires computationally expensive
matrix operations and involves a time-consuming model-selection procedure. We
propose a novel modification of the \emph{RDA} (\emph{GRDA}) classifier that
provides a straightforward interpration and employs a novel dimension-reduction
technique to enable a computationally efficient model-selection procedure, thus
overcoming the computational shortcomings of the \emph{RDA}
classifier. Moreover, we demonstrate with several high-dimensional microarray
data sets that our proposed classifier is competitive and often superior to
several recently proposed sparse and regularized classifiers in terms of
classification accuracy. Finally, we provide an implementation of the
\emph{GRDA} classifier in the {\tt sparsediscrim} R package, available on CRAN.

@@

In this paper we modify the \emph{RDA} classifier by first simplifying the
pooling matrix employed in the \emph{RDA} classifier. We construct a convex
combination of the covariance-matrix estimators employed in the \emph{LDA} and
\emph{QDA} classifiers and demonstrate that this simplification improves the
interpretation of the pooling operation of the \emph{RDA} classifier. The
\emph{LDA} classifier involves the assumption that the population covariance
matrices are equal, which is often too stringent in practice
\citep{Clemmensen:2011kr}. The \emph{GRDA} classifier allows us to relax the
linearity assumption by including unequal covariance matrices in the model. The
\emph{GRDA} classifier is then a compromise between the \emph{LDA} and
\emph{QDA} classifiers, so that the \emph{GRDA} classifier includes both the
\emph{LDA} and \emph{QDA} classifiers as special cases. We then introduce a more
general shrinkage term to the \emph{RDA} classifier to establish our \emph{GRDA}
classifier.

We show that the \emph{GRDA} classifier is invariant to any adjustments of the
eigenvalues corresponding to the null space of the pooled sample covariance
matrix. Applying a reasoning similar to \cite{Ye:2006tq}, we decompose the
\emph{RDA} classifier and show that the sum related to the null space of the
pooled sample covariance matrix can be discarded. Hence, no gain in
classification performance can be obtained by adjusting these zero
eigenvalues. This invariance property addresses the computational shortcomings
of the \emph{RDA} classifier and provides a substantial increase in the
computational efficiency of the \emph{RDA} classifier and its model
selection. In fact, we show that the inverse and determinants can be reduced to
calculations of $n_k \times n_k$ matrices, where $n_k$ is the sample size of the
$k$th class, $k = 1, \ldots, K$. This reduction in calculation is clearly
advantageous because $n_k$ is often quite small, say, 15--20, in microarray data
sets.

In addition to the improvement in computational performance, we demonstrate that
our \emph{GRDA} classifier is competitive with and often superior to several
recent classifiers designed for small-sample, high-dimensional data. In
particular, we compare our \emph{GRDA} classifier with the \emph{DLDA}
classifier from \cite{Dudoit:2002ev} and the recently proposed classifiers from
\cite{Guo:2007te}, \cite{Pang:2009ik}, \cite{Tong:2012hw}, and
\cite{Witten:2011kc}. We conduct our comparison of the classification accuracy
of the competing classifiers using four small-sample, high-dimensional
microarray data sets.

We have organized the remainder of the paper as follows. In Section 2 we review
the \emph{RDA} classifier from \cite{Friedman:1989tm} and discuss our
generalization in Section 3. In Section 4 we describe four small-sample,
high-dimensional microarray data sets and then demonstrate that our proposed
\emph{GRDA} classifier is often superior to the competing classifiers in terms
of classification accuracy. We then conclude with a brief discussion in Section
5.

@@

\section{The GRDA Classifier}

Here, we define our generalization of the \emph{RDA} classifier by first
simplifying the formulation of $\widehat{\bm \Sigma}_k(\lambda)$ in
\eqref{eq:sig-lambda} and demonstrating its clear interpretation as a linear
combination of the crossproducts of the training observations centered by their
respective class sample means. We define
\begin{align}
  \widehat{\bm\Sigma}_k(\lambda) = (1 - \lambda) \widehat{\bm\Sigma}_k + \lambda \widehat{\bm\Sigma}.\label{eq:sig-lambda-alternative}
\end{align}
as a convex combination of the MLEs of $\bm \Sigma_k$ and $\bm \Sigma$, where
$\lambda \in [0, 1]$ is the \emph{pooling} parameter. Thus, we avoid the convex
combination of $n_k$ and $N$ as defined by \cite{Friedman:1989tm}. By rewriting
\eqref{eq:sig-lambda-alternative} in terms of the observations $\bm x_i$, $i =
1, \ldots, N$, each centered by its class sample mean, we attain a clear
interpretation of $\widehat{\bm\Sigma}_k(\lambda)$. We have that
\begin{align}
	\widehat{\bm\Sigma}_k(\lambda) &= \widehat{\bm\Sigma}_k + \lambda (\widehat{\bm\Sigma} - \widehat{\bm\Sigma}_k) \nonumber\\
%	&= \widehat{\bm\Sigma}_k + \lambda \left(\sum_{k' = 1}^K \frac{n_k}{N} \widehat{\bm\Sigma}_{k'}  - \widehat{\bm\Sigma}_k \right) \nonumber\\
	&= \left( \frac{1 - \lambda}{n_k} + \frac{\lambda}{N} \right) \widehat{\bm\Sigma}_k +  \frac{\lambda}{N} \sum_{\substack{k' = 1\\k' \ne k}} n_{k'} \widehat{\bm\Sigma}_{k'} \nonumber \\
%	&= \left( \frac{1 - \lambda}{n_k} + \frac{\lambda}{N} \right)\sum_{i=1}^N I(y_i = k) \bm x_i \bm x_i' +  \frac{\lambda}{N} \sum_{i=1}^N I(y_i \ne k) \bm x_i \bm x_i' \nonumber \\
	&= \sum_{i=1}^N c_{ik}(\lambda) \bm x_i \bm x_i',\label{eq:sig-lambda-alternative2}
\end{align}
where $c_{ik}(\lambda) = \lambda N^{-1} + (1 - \lambda)n_k^{-1}I(y_i = k)$.
From \eqref{eq:sig-lambda-alternative2}, $\lambda$ weights the contribution of
each of the $N$ observations to estimating $\bm \Sigma_k$. That is, for $0 <
\lambda \le 1$, we estimate $\bm \Sigma_k$ with $N$ rather than $n_k$
observations. Below, we will show that the pooling operation is advantageous in
increasing the rank of each $\widehat{\bm\Sigma}_k(\lambda)$ from
rank$(\widehat{\bm\Sigma}_k)$ to rank$(\widehat{\bm\Sigma})$ for $0 < \lambda
\le 1$. Notice that if $\lambda = 0$, then the observations from the remaining
$K - 1$ classes do not contribute to the estimation of $\bm \Sigma_k$,
corresponding to $\widehat{\bm \Sigma}_k$. Furthermore, if $\lambda = 1$, the
weights in \eqref{eq:sig-lambda-alternative2} reduce to $1/N$, corresponding to
$\widehat{\bm\Sigma}$. In Figure \ref{fig:grda-contours} we demonstrate the
effect of $\lambda$ on the contours of five multivariate normal populations with
unequal covariance matrices. For brevity, when $\lambda = 1$, we define $\bm X =
[\sqrt{c_{1k}(1)} \bm x_1', \ldots, \sqrt{c_{Nk}(1)} \bm x_N']'$ such that
$\widehat{\bm \Sigma} = \bm X' \bm X$. Similarly, for $\lambda = 0$, we define
$\bm X_k = [\sqrt{c_{1k}(0)} \bm x_1', \ldots, \sqrt{c_{Nk}(0)} \bm x_N']'$ such
that $\widehat{\bm \Sigma}_k = \bm X_k' \bm X_k$, $k = 1, \ldots, K$.

\[ \left[\text{Figure \ref{fig:grda-contours} goes about here }\right]\]

As we have discussed, several eigenvalue adjustment methods have been proposed
to increase eigenvalues (approximately) equal to 0. Here, we define a more
general eigenvalue adjustment of \eqref{eq:sig-lambda-alternative} as
\begin{align}
	\tilde{\bm \Sigma}_k = \alpha_k \widehat{\bm \Sigma}_k(\lambda) + \gamma \bm I_p,\label{eq:grda-cov}
\end{align}
where $\alpha_k \ge 0$ and $\gamma \ge 0$ is an eigenvalue-shrinkage
constant. Thus, the \emph{GRDA} classifier is
\begin{align}
	D_{GRDA}(\bm x) = \argmin_{k}  (\bm x - \xbar_k)'\tilde{\bm\Sigma}_k^{+}(\bm x - \xbar_k)  + \ln |\tilde{\bm\Sigma}_k|. \label{eq:grda}
\end{align}
For $\gamma > 0$, $\tilde{\bm\Sigma}_k$ is nonsingular such that
$\tilde{\bm\Sigma}_k^{-1}$ can be substituted for $\tilde{\bm\Sigma}_k^{+}$ in
\eqref{eq:grda}. If $\gamma = 0$, we explicitly set $|\tilde{\bm\Sigma}_k|$
equal to the product of the positive eigenvalues of $\tilde{\bm\Sigma}_k$.

Next, we demonstrate that the \emph{GRDA} classifier lends itself to an
efficient calculation of \eqref{eq:grda} and a rapid model selection of the
pooling parameter $\lambda$. We decompose \eqref{eq:grda} into a sum of two
components, where the first summand consists of matrix operations applied to
low-dimensional matrices and the second summand corresponds to the null space of
$\widehat{\bm \Sigma}$ in \eqref{eq:pooled-cov}. We show that the matrix
operations performed on the null space of $\widehat{\bm \Sigma}$ yield constant
quadratic forms across all classes and can be omitted under mild regularity
conditions. For $p \gg N$, the constant component involves determinants and
inverses of high-dimensional matrices, and by ignoring these calculations, a
substantial reduction in computational costs is achieved relative to Friedman's
\emph{RDA} classifier in \eqref{eq:rda}. Furthermore, a byproduct is that
adjustments to the associated eigenvalues have no effect on
\eqref{eq:grda}. Lastly, we exploit the so-called fast singular value
decomposition (SVD) to construct the eigenvalue decomposition of $\widehat{\bm
  \Sigma}$, further reducing the computational costs of our proposed \emph{RDA}
classifier.

@@

Thus, by incorporating each $\bm x_i$ into the estimation of $\bm \Sigma_k$, we
increase the rank of $\widehat{\bm \Sigma}_k(\lambda)$ to $q$, $k = 1, \ldots,
K$ if the pooling parameter $\lambda \ne 0$. Next, we provide an essential
result that enables us to prove that, under mild regularity conditions,
\eqref{eq:grda} is invariant to adjustments to the eigenvalues of $\tilde{\bm
  \Sigma}_k$ corresponding to the null space of $\widehat{\bm \Sigma}$ for any
unlabeled observation $\bm x \in \mathbb{R}_{p \times 1}$.

@@

We now present our main result, where we decompose the \emph{GRDA} decision rule
and show that the component requiring the largest computational costs does not
contribute to the classification of an unlabeled observation performed using
Lemma \ref{lemma:RDA-constant-term}. Our main result reduces \eqref{eq:grda} to
an equivalent, more computationally efficient decision rule.

\begin{thm}
Let $\tilde{\bm \Sigma}_k$ and $\bm W_k$ be defined as in \eqref{eq:rda-matrix}
and \eqref{eq:Wk}, respectively, and let $\bm U_1$ be defined as above. Then,
the \emph{GRDA} decision rule in \eqref{eq:grda} is equivalent to
	\begin{align}
		D_{GRDA}(\bm x) &= \argmin_k  (\bm x - \xbar_k)' \bm U_1 \bm W_k^{-1} \bm U_1' (\bm x - \xbar_k) + \ln | \bm W_k |. \label{eq:rda-decomposed}
	\end{align}

@@

Thus far we have shown that the \emph{GRDA} classifier in
\eqref{eq:rda-decomposed} is invariant to the term $\bm U_2$, thus yielding a
substantial reduction in the computational costs. While the classifier is much
improved in terms of computational performance compared to the \emph{RDA}
classifier originally proposed by \cite{Friedman:1989tm}, we provide two
additional results that further improve the computational performance of our
\emph{GRDA} classifier. First, we demonstrate that the inverse and determinants
of $\bm W_k$ can be written so that they are performed on $n_k \times n_k$
matrices rather than the larger $q \times q$ matrices, resulting in further
improvements to the computational efficiency of the \emph{GRDA} classifier.

@@

In this section, we describe four high-dimensional microarray data sets and
compare our proposed \emph{GRDA} classifier with five recently proposed
classifiers designed for small-sample, high-dimensional data: the \emph{DLDA}
classifier from \cite{Dudoit:2002ev}, the shrunken centroids \emph{RDA}
classifier from \cite{Guo:2007te}, the shrinkage-based \emph{DLDA} classifier
from \cite{Pang:2009ik}, the shrinkage-mean-based DLDA classifier from
\cite{Tong:2012hw}, and the penalized \emph{LDA} classifier from
\cite{Witten:2011kc}. Below, we refer to each classifier by the first author's
surname. We evaluated the classification performance of each classifier by
randomly partitioning the data such that $2/3$ of the observations were
allocated as training data and the remaining $1/3$ of the observations were
allocated as a test data set. For each random partition, we applied the
feature-selection method from \cite{Dudoit:2002ev} to choose the $d$ features
having the largest ratio of their between-group to within-group sums of squares
from the training data. We then filtered the test data set with the
corresponding $d$ features. After training each classifier from the filtered
training data, we classified the filtered test data sets and computed the
proportion of mislabeled test observations to estimate the classification error
rate for each classifier. Repeating this process 1000 times, we computed the
average of the error-rate estimates for each classifier. We remark that our
simulation design avoids the selection bias discussed in
\cite{Ambroise:2002fa}. We did not include the \emph{RDA} classifier from
\cite{Friedman:1989tm} in our studies to avoid its excessive computations. In
Table \ref{tab:data-summary}, we summarize the attributes of each microarray
data set.

We used version 3.0.0 of the open-source statistical software {\tt R} for the
classification study. For each classifier's {\tt R} implementation, we used the
default settings with the exception that we explicitly set the prior
probabilities as equal. In our {\tt R} implementation of the \emph{GRDA}
classifier in the {\tt sparsediscrim} package, we set $\alpha_k = 1$, $k = 1,
\ldots, K$ for simplicity. We estimated $\lambda$ from a grid of 21 equidistant
candidate values between 0 and 1, inclusively. Similarly, we estimated $\gamma$
from a grid consisting of the values $10^{-1}, \ldots, 10^4$, and $10^5$. We
selected optimal estimates of $\lambda$ and $\gamma$ using 10-fold
cross-validation \citep{Hastie:2008dt}.

@@

The \emph{GRDA} classifier was superior in classification performance for the
majority of the simulations. For the Burczysnki, Shipp, and Singh data sets, the
\emph{GRDA} classifier outperformed its competitors for all values of $d$
examined. For the Nakayama data set, the \emph{GRDA} classifier was outperformed
by the Dudoit, Pang, and Tong classifiers for small values of $d$ but attained
classification rates superior to its competitors for larger values of
$d$. Furthermore, Figure \ref{fig:error-rates} emphasizes two strengths of the
\emph{GRDA} classifier: reliability of the classifier for small $d$ and
stability of the classifier for large $d$. In contrast to the other classifiers,
the \emph{GRDA} classifier exhibited less variability its misclassification
estimates for small values of $d$. Also, the \emph{GRDA} classifier yielded
superior classification accuracy as $d$ increased and was hardly effected by the
inclusion of more variables in the data sets. We attribute the apparent
stability of the error-rate estimates of the \emph{GRDA} classifier for large
values of $d$ to its novel dimension-reduction technique.

@@

Next, we examine the optimal estimates of the tuning parameters for the
\emph{GRDA} classifier. For brevity, we include only a graphical summary of the
optimal estimates of $\lambda$ for the Shipp data set in Figure
\ref{fig:optimal-shipp}. Similar figures for the other three data sets can be
found in the Supplementary Information. In Figure \ref{fig:optimal-shipp}, the
distribution of the optimal values of $\lambda$ selected through
cross-validation has a notable interesting bathtub shape, resembling a Beta
distribution with shape parameters $1/2$ and $1/2$. Specifically, the optimal
values of $\lambda$ are typically near 0 or near 1, corresponding to the
\emph{QDA} and \emph{LDA} classifiers, respectively. We noticed a similar
distribution of optimal values of $\lambda$ for the other three data sets.

>>>>

results$lambda <- as.numeric(results$lambda)
results$gamma <- as.numeric(results$gamma)
results$Dudoit <- as.numeric(results$Dudoit)
results$HDRDA <- as.numeric(results$HDRDA)
results$Guo <- as.numeric(results$Guo)
results$Pang <- as.numeric(results$Pang)
results$Tong <- as.numeric(results$Tong)
results$Witten <- as.numeric(results$Witten)

@@

# Summary of the optimal tuning-parameter estimates for each data set and number
# of variables selected, d.
hdrda_optimal <- subset(results, select = c(data_set, d, lambda, gamma))
hdrda_optimal$lambda <- factor(hdrda_optimal$lambda, levels = seq_lambda)
hdrda_optimal$gamma <- factor(hdrda_optimal$gamma, levels = seq_gamma)

hdrda_optimal <- ddply(hdrda_optimal, .(data_set, d), function(x) {
  x <- subset(x, select = c(lambda, gamma))
  melt(table(x))
})

@@

A reliable supervised classification of tumors from gene expression microarray
data is essential for successful diagnosis and treatment of diseases such as
cancer; however, the analysis of microarray data often presents difficult
challenges because the number of gene expression levels is typically much
greater than the sample size. Consequently, several sparse and regularized
classifiers have been proposed, often with strict covariance assumptions or
eigenvalue adjustments to ensure that the classifiers are well-posed. Here, we
consider an improved version of the regularized discriminant analysis
(\emph{RDA}) classifier popularized by \cite{Friedman:1989tm}. That is, we
propose the high-dimensional \emph{RDA} (\emph{HDRDA}) classifier, a novel
modification of the \emph{RDA} classifier, that provides a straightforward
interpretation and employs a new dimension-reduction technique to enable a
computationally efficient model-selection procedure, thus overcoming the
computational shortcomings of the \emph{RDA} classifier for high-dimensional
data. Moreover, we demonstrate with several high-dimensional microarray data
sets that our proposed classifier is competitive and often superior to several
recently proposed sparse and regularized classifiers in terms of classification
accuracy. Finally, we provide an implementation of the \emph{HDRDA} classifier in
the {\tt sparsediscrim} R package, available on CRAN.

@@

In this paper we modify the \emph{RDA} classifier by first simplifying the
pooling matrix employed in the \emph{RDA} classifier. We construct a convex
combination of the covariance-matrix estimators employed in the \emph{LDA} and
\emph{QDA} classifiers and demonstrate that this simplification improves the
interpretation of the pooling operation of the \emph{RDA} classifier. The
\emph{LDA} classifier involves the assumption that the population covariance
matrices are equal, which is often too stringent in practice
\citep{Clemmensen:2011kr}. The \emph{HDRDA} classifier allows us to relax the
linearity assumption by including unequal covariance matrices in the model. The
\emph{HDRDA} classifier is then a compromise between the \emph{LDA} and
\emph{QDA} classifiers, so that the \emph{HDRDA} classifier includes both the
\emph{LDA} and \emph{QDA} classifiers as special cases. We then introduce a more
general shrinkage term to the \emph{RDA} classifier to establish our \emph{HDRDA}
classifier.

We show that the \emph{HDRDA} classifier is invariant to any adjustments of the
eigenvalues corresponding to the null space of the pooled sample covariance
matrix. Applying a reasoning similar to \cite{Ye:2006tq}, we decompose the
\emph{RDA} classifier and show that the sum related to the null space of the
pooled sample covariance matrix can be discarded. Hence, no gain in
classification performance can be obtained by adjusting these zero
eigenvalues. This invariance property addresses the computational shortcomings
of the \emph{RDA} classifier and provides a substantial increase in the
computational efficiency of the \emph{RDA} classifier and its model
selection. In fact, we show that the inverse and determinants can be reduced to
calculations of $n_k \times n_k$ matrices, where $n_k$ is the sample size of the
$k$th class, $k = 1, \ldots, K$. This reduction in calculation is clearly
advantageous because $n_k$ is often quite small, say, 15--20, in microarray data
sets.

In addition to the improvement in computational performance, we demonstrate that
our \emph{HDRDA} classifier is competitive with and often superior to several
recent classifiers designed for small-sample, high-dimensional data. In
particular, we compare our \emph{HDRDA} classifier with the \emph{DLDA}
classifier from \cite{Dudoit:2002ev} and the recently proposed classifiers from
\cite{Guo:2007te}, \cite{Pang:2009ik}, \cite{Tong:2012hw}, and
\cite{Witten:2011kc}. We conduct our comparison of the classification accuracy
of the competing classifiers using four small-sample, high-dimensional
microarray data sets.

We have organized the remainder of the paper as follows. In Section 2 we review
the \emph{RDA} classifier from \cite{Friedman:1989tm} and discuss our
generalization in Section 3. In Section 4 we describe four small-sample,
high-dimensional microarray data sets and then demonstrate that our proposed
\emph{HDRDA} classifier is often superior to the competing classifiers in terms
of classification accuracy. We then conclude with a brief discussion in Section
5.

@@

\section{The HDRDA Classifier}

Here, we define our generalization of the \emph{RDA} classifier by first
simplifying the formulation of $\widehat{\bm \Sigma}_k(\lambda)$ in
\eqref{eq:sig-lambda} and demonstrating its clear interpretation as a linear
combination of the crossproducts of the training observations centered by their
respective class sample means. We define
\begin{align}
  \widehat{\bm\Sigma}_k(\lambda) = (1 - \lambda) \widehat{\bm\Sigma}_k + \lambda \widehat{\bm\Sigma}.\label{eq:sig-lambda-alternative}
\end{align}
as a convex combination of the MLEs of $\bm \Sigma_k$ and $\bm \Sigma$, where
$\lambda \in [0, 1]$ is the \emph{pooling} parameter. Thus, we avoid the convex
combination of $n_k$ and $N$ as defined by \cite{Friedman:1989tm}. By rewriting
\eqref{eq:sig-lambda-alternative} in terms of the observations $\bm x_i$, $i =
1, \ldots, N$, each centered by its class sample mean, we attain a clear
interpretation of $\widehat{\bm\Sigma}_k(\lambda)$. We have that
\begin{align}
	\widehat{\bm\Sigma}_k(\lambda) &= \widehat{\bm\Sigma}_k + \lambda (\widehat{\bm\Sigma} - \widehat{\bm\Sigma}_k) \nonumber\\
%	&= \widehat{\bm\Sigma}_k + \lambda \left(\sum_{k' = 1}^K \frac{n_k}{N} \widehat{\bm\Sigma}_{k'}  - \widehat{\bm\Sigma}_k \right) \nonumber\\
	&= \left( \frac{1 - \lambda}{n_k} + \frac{\lambda}{N} \right) \widehat{\bm\Sigma}_k +  \frac{\lambda}{N} \sum_{\substack{k' = 1\\k' \ne k}} n_{k'} \widehat{\bm\Sigma}_{k'} \nonumber \\
%	&= \left( \frac{1 - \lambda}{n_k} + \frac{\lambda}{N} \right)\sum_{i=1}^N I(y_i = k) \bm x_i \bm x_i' +  \frac{\lambda}{N} \sum_{i=1}^N I(y_i \ne k) \bm x_i \bm x_i' \nonumber \\
	&= \sum_{i=1}^N c_{ik}(\lambda) \bm x_i \bm x_i',\label{eq:sig-lambda-alternative2}
\end{align}
where $c_{ik}(\lambda) = \lambda N^{-1} + (1 - \lambda)n_k^{-1}I(y_i = k)$.
From \eqref{eq:sig-lambda-alternative2}, $\lambda$ weights the contribution of
each of the $N$ observations to estimating $\bm \Sigma_k$. That is, for $0 <
\lambda \le 1$, we estimate $\bm \Sigma_k$ with $N$ rather than $n_k$
observations. Below, we will show that the pooling operation is advantageous in
increasing the rank of each $\widehat{\bm\Sigma}_k(\lambda)$ from
rank$(\widehat{\bm\Sigma}_k)$ to rank$(\widehat{\bm\Sigma})$ for $0 < \lambda
\le 1$. Notice that if $\lambda = 0$, then the observations from the remaining
$K - 1$ classes do not contribute to the estimation of $\bm \Sigma_k$,
corresponding to $\widehat{\bm \Sigma}_k$. Furthermore, if $\lambda = 1$, the
weights in \eqref{eq:sig-lambda-alternative2} reduce to $1/N$, corresponding to
$\widehat{\bm\Sigma}$. For brevity, when $\lambda = 1$, we define $\bm X =
[\sqrt{c_{1k}(1)} \bm x_1', \ldots, \sqrt{c_{Nk}(1)} \bm x_N']'$ such that
$\widehat{\bm \Sigma} = \bm X' \bm X$. Similarly, for $\lambda = 0$, we define
$\bm X_k = [\sqrt{c_{1k}(0)} \bm x_1', \ldots, \sqrt{c_{Nk}(0)} \bm x_N']'$ such
that $\widehat{\bm \Sigma}_k = \bm X_k' \bm X_k$, $k = 1, \ldots, K$. In Figure
\ref{fig:hdrda-contours} we demonstrate the effect of $\lambda$ on the contours
of five multivariate normal populations with unequal covariance matrices.

\[ \left[\text{Figure \ref{fig:hdrda-contours} goes about here }\right]\]

As we have discussed, several eigenvalue adjustment methods have been proposed
to increase eigenvalues (approximately) equal to 0. Here, we define the
eigenvalue adjustment of \eqref{eq:sig-lambda-alternative} as
\begin{align}
	\tilde{\bm \Sigma}_k = \alpha_k \widehat{\bm \Sigma}_k(\lambda) + \gamma \bm I_p,\label{eq:hdrda-cov}
\end{align}
where $\alpha_k \ge 0$ and $\gamma \ge 0$ is an eigenvalue-shrinkage
constant. The choice of $\alpha_k$ allows for a flexible formulation of
covariance-matrix estimators. For instance, if $\alpha_k = 1$, $k = 1, \ldots,
K$, then \eqref{eq:hdrda-cov} resembles \eqref{eq:ridge-estimator}. Similarly,
if $\alpha_k = 1 - \gamma$, then \eqref{eq:hdrda-cov} has a form similar to
\eqref{eq:sig-rda}. Now, substituting \eqref{eq:hdrda-cov} into \eqref{eq:qda},
we define the \emph{HDRDA} classifier as
\begin{align}
	D_{HDRDA}(\bm x) = \argmin_{k}  (\bm x - \xbar_k)'\tilde{\bm\Sigma}_k^{+}(\bm x - \xbar_k)  + \ln |\tilde{\bm\Sigma}_k|. \label{eq:hdrda}
\end{align}
For $\gamma > 0$, $\tilde{\bm\Sigma}_k$ is nonsingular such that
$\tilde{\bm\Sigma}_k^{-1}$ can be substituted for $\tilde{\bm\Sigma}_k^{+}$ in
\eqref{eq:hdrda}. If $\gamma = 0$, we explicitly set $|\tilde{\bm\Sigma}_k|$
equal to the product of the positive eigenvalues of $\tilde{\bm\Sigma}_k$.

Next, we demonstrate that the \emph{HDRDA} classifier lends itself to an
efficient calculation of \eqref{eq:hdrda} and a rapid model selection of the
pooling parameter $\lambda$. We decompose \eqref{eq:hdrda} into a sum of two
components, where the first summand consists of matrix operations applied to
low-dimensional matrices and the second summand corresponds to the null space of
$\widehat{\bm \Sigma}$ in \eqref{eq:pooled-cov}. We show that the matrix
operations performed on the null space of $\widehat{\bm \Sigma}$ yield constant
quadratic forms across all classes and can be omitted under mild regularity
conditions. For $p \gg N$, the constant component involves determinants and
inverses of high-dimensional matrices, and by ignoring these calculations, a
substantial reduction in computational costs is achieved relative to Friedman's
\emph{RDA} classifier in \eqref{eq:rda}. Furthermore, a byproduct is that
adjustments to the associated eigenvalues have no effect on
\eqref{eq:hdrda}. Lastly, we exploit the so-called fast singular value
decomposition (SVD) to construct the eigenvalue decomposition of $\widehat{\bm
  \Sigma}$, further reducing the computational costs of our proposed \emph{RDA}
classifier.

@@

Thus, by incorporating each $\bm x_i$ into the estimation of $\bm \Sigma_k$, we
increase the rank of $\widehat{\bm \Sigma}_k(\lambda)$ to $q$ if $\lambda \ne
0$. Next, we provide an essential result that enables us to prove that, under
mild regularity conditions, \eqref{eq:hdrda} is invariant to adjustments to the
eigenvalues of $\tilde{\bm \Sigma}_k$ corresponding to the null space of
$\widehat{\bm \Sigma}$ for any unlabeled observation $\bm x \in \mathbb{R}_{p
  \times 1}$.

@@

We now present our main result, where we decompose the \emph{HDRDA} decision rule
and show that the component requiring the largest computational costs does not
contribute to the classification of an unlabeled observation performed using
Lemma \ref{lemma:RDA-constant-term}. Our main result reduces \eqref{eq:hdrda} to
an equivalent, more computationally efficient and more stable decision rule.

\begin{thm}
Let $\tilde{\bm \Sigma}_k$ and $\bm W_k$ be defined as in \eqref{eq:rda-matrix}
and \eqref{eq:Wk}, respectively, and let $\bm U_1$ be defined as above. Then,
the \emph{HDRDA} decision rule in \eqref{eq:hdrda} is equivalent to
	\begin{align}
		D_{HDRDA}(\bm x) &= \argmin_k  (\bm x - \xbar_k)' \bm U_1 \bm W_k^{-1} \bm U_1' (\bm x - \xbar_k) + \ln | \bm W_k |. \label{eq:rda-decomposed}
	\end{align}

@@

Thus far we have shown that the \emph{HDRDA} classifier in
\eqref{eq:rda-decomposed} is invariant to the term $\bm U_2$, thus yielding a
substantial reduction in the computational costs. While the classifier is much
improved in terms of computational performance compared to the \emph{RDA}
classifier originally proposed by \cite{Friedman:1989tm}, we provide two
additional results that further improve the computational performance of our
\emph{HDRDA} classifier. First, we demonstrate that the inverse and determinants
of $\bm W_k$ can be written so that they are performed on $n_k \times n_k$
matrices rather than the larger $q \times q$ matrices, resulting in further
improvements to the computational efficiency of the \emph{HDRDA} classifier.

@@

In this section, we describe four high-dimensional microarray data sets and
compare our proposed \emph{HDRDA} classifier with five recently proposed
classifiers designed for small-sample, high-dimensional data: the \emph{DLDA}
classifier from \cite{Dudoit:2002ev}, the shrunken centroids \emph{RDA}
classifier from \cite{Guo:2007te}, the shrinkage-based \emph{DLDA} classifier
from \cite{Pang:2009ik}, the shrinkage-mean-based DLDA classifier from
\cite{Tong:2012hw}, and the penalized \emph{LDA} classifier from
\cite{Witten:2011kc}. Below, we refer to each classifier by the first author's
surname. We evaluated the classification performance of each classifier by
randomly partitioning the data such that $2/3$ of the observations were
allocated as training data and the remaining $1/3$ of the observations were
allocated as a test data set. For each random partition, we applied the
feature-selection method from \cite{Dudoit:2002ev} to choose the $d$ features
having the largest ratio of their between-group to within-group sums of squares
from the training data. We then filtered the test data set with the
corresponding $d$ features. After training each classifier from the filtered
training data, we classified the filtered test data sets and computed the
proportion of mislabeled test observations to estimate the classification error
rate for each classifier. Repeating this process 1000 times, we computed the
average of the error-rate estimates for each classifier. We remark that our
simulation design avoids the selection bias discussed in
\cite{Ambroise:2002fa}. We did not include the \emph{RDA} classifier from
\cite{Friedman:1989tm} in our studies to avoid its excessive computations. In
Table \ref{tab:data-summary}, we summarize the attributes of each microarray
data set.

We used version 3.0.0 of the open-source statistical software {\tt R} for the
classification study. For each classifier's {\tt R} implementation, we used the
default settings with the exception that we explicitly set the prior
probabilities as equal. In our {\tt R} implementation of the \emph{HDRDA}
classifier in the {\tt sparsediscrim} package, we set $\alpha_k = 1$, $k = 1,
\ldots, K$, so that the covariance-matrix estimator \eqref{eq:hdrda-cov}
employed in the \emph{HDRDA} classifier has a form resembling
\eqref{eq:ridge-estimator}. We estimated $\lambda$ from a grid of 21 equidistant
candidate values between 0 and 1, inclusively. Similarly, we estimated $\gamma$
from a grid consisting of the values $10^{-1}, \ldots, 10^4$, and $10^5$. We
selected optimal estimates of $\lambda$ and $\gamma$ using 10-fold
cross-validation \citep{Hastie:2008dt}.

@@

The \emph{HDRDA} classifier was superior in classification performance for the
majority of the simulations. For the Burczysnki, Shipp, and Singh data sets, the
\emph{HDRDA} classifier outperformed its competitors for all values of $d$
examined. For the Nakayama data set, the \emph{HDRDA} classifier was
outperformed by the Dudoit, Pang, and Tong classifiers for small values of $d$
but attained classification rates superior to its competitors for larger values
of $d$. Furthermore, Figure \ref{fig:error-rates} emphasizes two strengths of
the \emph{HDRDA} classifier: reliability of the classifier for small $d$ and
stability of the classifier for large $d$. In contrast to the other classifiers,
the \emph{HDRDA} classifier exhibited less variability in its misclassification
estimates for small values of $d$. Also, the \emph{HDRDA} classifier yielded
superior classification accuracy as $d$ increased and was hardly effected by the
inclusion of more variables in the data sets. We attribute the apparent
stability of the error-rate estimates of the \emph{HDRDA} classifier for large
values of $d$ to its novel dimension-reduction technique.

@@

Next, we examine the optimal estimates of the tuning parameters for the
\emph{HDRDA} classifier. For brevity, we include only a graphical summary of the
optimal estimates of $\lambda$ for the Shipp data set in Figure
\ref{fig:optimal-shipp}. Similar figures for the other three data sets can be
found in the Supplementary Information. In Figure \ref{fig:optimal-shipp}, the
distribution of the optimal values of $\lambda$ selected through
cross-validation has a notable bathtub shape, resembling a Beta distribution
with both shape parameters less than unity. Specifically, the optimal values of
$\lambda$ are typically near 0 or near 1, corresponding to the \emph{QDA} and
\emph{LDA} classifiers, respectively. We noticed a similar distribution of
optimal values of $\lambda$ for the other three data sets.
